# Variable Pre-Processing

Each code block should refer to the production of one dataset. Datasets should be saved in `data/variables`.

## Initialize Packages
```{r warning=FALSE}
here::i_am("code/variable-preprocessing.Rmd")

# Initialize packages
packages <- c("tidyverse", "caret","ggplot2", "modelr", "boot", "here", "sf", "raster", "prism", "exactextractr", "devtools")

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))

#Github packages
remotes::install_github("mikejohnson51/AOI") 
remotes::install_github("mikejohnson51/climateR")
library(AOI)
library(climateR)



set.seed(313)

prism_set_dl_dir(here("data/PRISM"))
```


## GW Basin Precipitation
Data source: PRISM
Timeframe: 1901-2022
Unit: mm

```{r}
#download PRISM data
get_prism_annual("ppt", years = 1901:2022, keepZip = FALSE)

#Format to pull produce raster:
ppt_2013<- pd_to_file(prism_archive_subset("ppt", "annual", years = 2013))
ppt_2013_rast <- raster(ppt_2013)

#GW Basins shapefile
GW_shape <- shapefile(here("data/GW_shapefile/Groundwater_Basin.shp"))

#set CRS to the same
GW_shape <- spTransform(GW_shape, crs(ppt_2013_rast))


#Assign annual precip to each GW Basin for period of record
GW_precip <- data.frame(HUC = GW_shape$BASIN_NAME) #initialize
count <- 1 #initialize

for(i in 1901:2022){ #period of record
  count <- count+1
  
  rast_file <- pd_to_file(prism_archive_subset("ppt", "annual", years = i)) #read raster filename
  tmp_rast <- raster(rast_file) #create raster
  
  #produce HUC means for that raster
  tmp_mean <- exact_extract(tmp_rast, GW_shape, fun = "mean")
  
  GW_precip[,count] <- round(tmp_mean,2) #input means to dataframe
  colnames(GW_precip)[count] <- c(as.character(i)) #rename column
}

# write_csv(GW_precip, here("data/variables/GW_Precipitation_ANNUAL.csv"))
```

## GW Basin Temperature
Data source: PRISM
Timeframe: 1901-2022
Unit: degrees C

```{r}
#download PRISM data
get_prism_annual("tmean", years = 1901:2022, keepZip = FALSE)

#Format to pull produce raster:
ppt_2013<- pd_to_file(prism_archive_subset("ppt", "annual", years = 2013))
ppt_2013_rast <- raster(ppt_2013)

#GW Basins shapefile
GW_shape <- shapefile(here("data/GW_shapefile/Groundwater_Basin.shp"))

#set CRS to the same
GW_shape <- spTransform(GW_shape, crs(ppt_2013_rast))


#Assign annual precip to each GWBasins for period of record
GW_temp <- data.frame(HUC = GW_shape$BASIN_NAME) #initialize
count <- 1 #initialize

for(i in 1901:2022){ #period of record
  count <- count+1
  
  rast_file <- pd_to_file(prism_archive_subset("tmean", "annual", years = i)) #read raster filename
  tmp_rast <- raster(rast_file) #create raster
  
  #produce HUC means for that raster
  tmp_mean <- exact_extract(tmp_rast, GW_shape, fun = "mean")
  
  GW_temp[,count] <- round(tmp_mean,2) #input means to dataframe
  colnames(GW_temp)[count] <- c(as.character(i)) #rename column
}

write_csv(GW_temp, here("data/variables/GW_Temperature_ANNUAL.csv"))
```

## GW Basin ET
Data source: TerraClimate
Timeframe: 1958-2022 (years before 1958 replaced with basin mean over period of record)
Unit: mm

```{r}
AZ <- aoi_get(state = "AZ")
GWBasins <- st_sf(st_read(here("data/GW_shapefile/Groundwater_Basin.shp")))

GW_annualET <- data_frame(GWBasin = GWBasins$BASIN_NAME) #dataframe with HUC8 numbers in order

for (year in 1958:2022){
  
  d <- getTerraClim(AZ, #shape of AZ
                  varname = "aet", #actual ET
                  startDate = paste(year,"01","01",sep = "-"), 
                  endDate= paste(year,"12","31", sep = "-"))
  
  
  #calculate mean annual ET
  summed_values <- terra::app(d$aet, sum)

  z <- exact_extract(summed_values, GWBasins, fun = "mean", append_cols = "BASIN_NAME")
  
  z <- z %>%
    rename_with(~paste0(as.character(year)), mean)
  
  GW_annualET <- cbind(GW_annualET,z[2])
}

#Produce mean annual ET for each basin across available period of record of terraClimate data (1958-2022) to be the ET values pre-1958

GWMean <- rowMeans(as.matrix(GW_annualET[,2:66]))
GW_annualET <- cbind(GW_annualET,GWMean)

# Make columns for 1901-1957
years <- data.frame(matrix(ncol = length(1901:1957), nrow = nrow(GW_annualET)))
colnames(years) <- as.character(1901:1957)

GW_annualET <- cbind(years, GW_annualET)

# Replace all NA from 1901-1957 with mean basin ET
for (i in 1901:1957){
  GW_annualET[, as.character(i)] <- GW_annualET[,124]
}

# Reorder columns
GW_annualET <- GW_annualET[,c(58,1:57,59:123)]

#write_csv(GW_annualET, here("data/variables/GW_ET_ANNUAL.csv"))
```


## GW Basin USGS Streamgage-keyed data
Data source: Various
Timeframe: 1901-2022
Goal: Assign precipitation, temperature, and ET data to each year/GWBasin

```{r}
GaugeData <- read_csv(here("data/GWBasins_GaugeData_FULL.csv"))

for(i in 1:nrow(GaugeData)){
  yr <- GaugeData$YEAR[i]
  yr_index <- which(colnames(GW_Precipitation_ANNUAL) == yr)
  
  basin <- GaugeData$GWBASIN[i]
  basin_index <- which(GW_Precipitation_ANNUAL$GWBasin == basin)
  
  #Precipitation
  GaugeData[i,7] <- as.numeric(GW_Precipitation_ANNUAL[basin_index,yr_index])
  
  #Temperature
  GaugeData[i,8] <- as.numeric(GW_Temperature_ANNUAL[basin_index,yr_index])
  
  #ET
  GaugeData[i,12] <- as.numeric(GW_ET_ANNUAL[basin_index,yr_index])
}

#write_csv(GaugeData, here("data/GWBasins_GaugeData_FULL.csv"))
```
